---
title: "Notes on simulation"
format:
  html:
    toc: true
    embed-resources: true
---

```{r}
#| echo: false
#| message: false

library(here)
library(igraph)
library(ggplot2)
library(RColorBrewer)

#------------------------------------------------------------------#
# Plotting functions
#------------------------------------------------------------------#
plot_network_geo <- function(sim_obj, main = NULL) {
  Y <- sim_obj$Y
  x <- sim_obj$x
  z <- sim_obj$partition

  g <- igraph::graph_from_adjacency_matrix(
    Y,
    mode = "undirected",
    diag = FALSE
  )

  layout_xy <- as.matrix(x)   # use (x1, x2) as coordinates

  K    <- length(unique(z))
  pal  <- c(1, 2, 3,4)
  vcol <- pal[z]

  if (is.null(main)) {
    main <- paste0("Network with geographic layout (cov_dep = '",
                   sim_obj$cov_dep, "')")
  }

  plot(
    g,
    layout         = layout_xy,
    vertex.color   = vcol,
    vertex.size    = 6,
    vertex.label   = NA,
    edge.arrow.mode = 0,
    main           = main
  )
}


plot_adj_matrix_gg <- function(sim_obj) {
  Y  <- sim_obj$Y
  z0 <- sim_obj$partition
  V  <- length(z0)
  
  # remove self-loops
  diag(Y) <- 0
  
  #------------------------------------------#
  # Order nodes by true cluster to show blocks
  #------------------------------------------#
  ord   <- order(z0)
  Y_ord <- Y[ord, ord]
  z_ord <- z0[ord]
  
  #------------------------------------------#
  # Build long data.frame in base R
  #------------------------------------------#
  # indices
  grid_idx <- expand.grid(
    row = seq_len(V),
    col = seq_len(V)
  )
  
  # adjacency values (row-major)
  grid_idx$value <- as.vector(Y_ord)
  
  # factors for plotting:
  # flip row so cluster 1 is at bottom (heatmap style)
  grid_idx$row_f <- factor(grid_idx$row,
                           levels = rev(seq_len(V)))
  grid_idx$col_f <- factor(grid_idx$col,
                           levels = seq_len(V))
  
  # cluster label per row (after ordering)
  grid_idx$cluster <- factor(z_ord[grid_idx$row])
  
  #------------------------------------------#
  # Colors
  #------------------------------------------#
  adj_pal <- colorRampPalette(
    brewer.pal(9, "Greys")[c(1, 9)]
  )(50)
  
  #------------------------------------------#
  # Plot
  #------------------------------------------#
  ggplot(grid_idx, aes(x = col_f, y = row_f, fill = value)) +
    geom_tile() +
    scale_fill_gradientn(colors = adj_pal, guide = "none") +
    coord_fixed() +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x  = element_blank(),
      axis.text.y  = element_blank(),
      axis.title   = element_blank(),
      panel.grid   = element_blank()
    ) +
    ggtitle("Adjacency matrix")
}


plot_psm_gg <- function(psm, title = "") {
  V <- nrow(psm)
  
  # long data.frame in base R
  grid_idx <- expand.grid(
    row = seq_len(V),
    col = seq_len(V)
  )
  grid_idx$value <- as.vector(psm)
  
  # flip rows so origin is bottom-left like a matrix
  grid_idx$row_f <- factor(grid_idx$row, levels = rev(seq_len(V)))
  grid_idx$col_f <- factor(grid_idx$col, levels = seq_len(V))
  
  pal <- colorRampPalette(brewer.pal(9, "Blues"))(50)
  
  ggplot(grid_idx, aes(x = col_f, y = row_f, fill = value)) +
    geom_tile() +
    scale_fill_gradientn(colors = pal, limits = c(0, 1)) +
    coord_fixed() +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x  = element_blank(),
      axis.text.y  = element_blank(),
      axis.title   = element_blank(),
      panel.grid   = element_blank(),
      plot.title   = element_text(hjust = 0.5)
    ) +
    ggtitle(title)
}
```


## Simulation design

We consider a transportation--network setting in which each node is characterized by two covariates that can be interpreted as a geographic location (longitude and latitude). The network contains $150$ nodes partitioned into four clusters of sizes $(50, 40, 30, 30)$. Edges are generated from a stochastic block model with higher within--cluster than between--cluster connection probabilities (0.4 vs 0.2).  We examine three scenarios, which determine how informative the geographic coordinates are about the true clusters: (i) \emph{no covariate effect}, where both coordinates are independent of the clusters; (ii) \emph{single--coordinate effect}, where only the first coordinate varies across clusters and the second is noise; and (iii) \emph{two--coordinate effect}, where both coordinates differ across clusters and thus jointly reveal the underlying community structure. 


```{r}
#| layout-ncol: 2
sim_none  <- readRDS(here("simulation", "data", "binarySBM_none.rds"))
sim_one  <- readRDS(here("simulation", "data", "binarySBM_one.rds"))
sim_both  <- readRDS(here("simulation", "data", "binarySBM_both.rds"))


plot_adj_matrix_gg(sim_none)
plot_network_geo(sim_none)
plot_network_geo(sim_one)
plot_network_geo(sim_both)
```

**Note**: for this first simualtion I used just scenario none and one and included only one covariate in the model

For the simulation study we run the ESBM gibbs with the following settings 

- Block probabilities:
  - Each entry of the block matrix has a $\text{Beta}(a,b)$ prior with
    $a = 1$, $b = 1$ (uniform on $(0,1)$).
- Random partition prior:
  - We use a Dirichlet–multinomial (DM) prior on cluster allocations, with concentration parameter $\beta_{\text{DM}} = 1$ and $H_{\text{DM}} = 4$ (same as simulation for now)
- Similary function for the covariante is a Gaussian pseduo model

$$
  x_i \mid \mu_h \sim N(\mu_h, 1), \qquad
  \mu_h \sim N(m_0, s_0^2),
$$

with $(m_0, s_0^2) = (0, 1)$, 
    
- $alpha$ is fixed and not sampled at values $(0,0.5, 1, 2, 4)$

We run the MCMC for 10000 iterations

### Results - one covariate